Fitness tracking reveals task-specific associations between memory, mental health, and physical activity

### 2022-07-14 paper/main.tex
updating paper

\section*{Introduction}
Engaging in physical activity (exercise) can improve physical fitness
by increasing muscle strength \citep{RogeEvan93, Lind79, CranEtal13,
  Knut07}, bone density~\citep{ChilEtal12, BassRams94, LaynNels99},
cardiovascular performance~\citep{MaioEtal00, PollEtal00}, lung
capacity~\citep[][although see \cite{RomaEtal16}]{LazoEtal16}, and
endurance~\citep{WilmKnut03}.  Physical activity can also improve mental
health~\citep{Ragl90, MikkEtal17, TaylEtal85, DeslEtal09, Call04,
  PaluSchw00, BassSuzu17} and cognitive
performance~\citep{ChanEtal12b, BrisEtal02, EtniEtal06, BassSuzu17}.

@@

In the present study, we hypothesize that (a) different exercise
regimens will have different, quantifiable impacts on cognitive
performance and mental health, and that (b) these impacts will be
consistant across individuals.  To this end, we collected a year of
real-world fitness tracking data from each of 113 participants.  We
then asked each participant to fill out a brief survey in which they
self-evaluated and self-reported several aspects of their mental
health.  Finally, we ran each participant through a battery of memory
tasks, which we used to evaluate their memory performance along
several dimensions.  We searched the data for potential associations
between memory, mental health, and physical activity.

@@

We wondered about potential links between the different aspects of
participants' data.  For example, if people who exercised in a
particular way also tended to perform better on a given memory task,
this could suggest that either (a) some property intrinsic to
participants who exercised in a particular way might also affect their
memory performance on the given task, and/or (b) particular physical activity
behaviors could have a causal impact on memory performance.  We
carried out an exploratory analysis whereby we used a bootstrap-based
approach to identify reliable correlations between different aspects
of memory performance (Fig.~\behBehCorr), different aspects of fitness
(Fig.~\fitFitCorr), different demographic attributes
(Fig.~\demDemCorr), and correlations between memory performance,
fitness information, and demographic attributes (Fig.~\allCorr).
Specifically, we sought to identify correlations that were present in
the same direction (i.e., positive or negative) across different
subsets of participants.  For each test, we report the average correlation (taken
across 10,000 subsets of participants, chosen with replacement) and
an associated two-tailed $p$-value, estimated as
\[
p = 2 \times (1 - q),
\]
where $q$ is the proportion of those
10,000 subsets that exhibited correlations in the same direction  (see
\textit{Exploratory correlation analyses}).  When all 10,000 randomly
chosen subsets of participants exhibited correlations in the same
direction (i.e., all positive correlations or all negative
correlations), we report the $p$-value as $p < 0.0001$.

@@

\section*{Discussion}
After collecting a year's worth of fitness-tracking data from each of 113
participants, we ran each participant in a battery of memory tasks and
had them fill out a series of demographic and mental health-related
questions.  We found that the associations between fitness-related
activities, memory performance, and mental health were heterogeneous.
Our results suggest that engagement in particular physical
activities (e.g., differing in time relative to the
test day, intensity, duration, etc.) is also reflected in participants'
memory performance patterns across tasks and in participants' mental
health attributes.

>>>>

\section*{Introduction}
Engaging in physical activity (exercise) can improve physical fitness
by increasing muscle strength \citep{RogeEvan93, Lind79, CranEtal13,
  Knut07}, bone density~\citep{ChilEtal12, BassRams94, LaynNels99},
cardiovascular performance~\citep{MaioEtal00, PollEtal00}, lung
capacity~\citep[][although see \cite{RomaEtal16}]{LazoEtal16}, and
endurance~\citep{WilmKnut03}.  Physical activity can also improve mental
health~\citep{Ragl90, MikkEtal17, TaylEtal85, DeslEtal09, Call04,
  PaluSchw00, BassSuzu17, MorrEtal18, GordEtal17, MorrEtal22, HerrEtal10} and cognitive
performance~\citep{ChanEtal12b, BrisEtal02, EtniEtal06, BassSuzu17}.

@@

In the present study, we hypothesize that (a) different intensities of physical activity will have different, quantifiable impacts on cognitive
performance and mental health, and that (b) these impacts will be
consistant across individuals.  To this end, we collected a year of
real-world fitness tracking data from each of 113 participants.  We
then asked each participant to fill out a brief survey in which they
self-evaluated and self-reported several aspects of their mental
health.  Finally, we ran each participant through a battery of memory
tasks, which we used to evaluate their memory performance along
several dimensions.  We searched the data for potential associations
between memory, mental health, and physical activity.

@@

We wondered about potential links between the different aspects of
participants' data.  For example, if people who engaged in particular intensities of physical activity also tended to perform better on a given memory task,
this could suggest that either (a) some property intrinsic to
participants who exercised in a particular way might also affect their
memory performance on the given task, and/or (b) particular physical activity
behaviors could have a causal impact on memory performance.  We
carried out an exploratory analysis whereby we used a bootstrap-based
approach to identify reliable correlations between different aspects
of memory performance (Fig.~\behBehCorr), different aspects of fitness
(Fig.~\fitFitCorr), different demographic attributes
(Fig.~\demDemCorr), and correlations between memory performance,
fitness information, and demographic attributes (Fig.~\allCorr).
Specifically, we sought to identify correlations that were present in
the same direction (i.e., positive or negative) across different
subsets of participants.  For each test, we report the average correlation (taken
across 10,000 subsets of participants, chosen with replacement) and
an associated two-tailed $p$-value, estimated as
\[
p = 2 \times (1 - q),
\]
where $q$ is the proportion of those
10,000 subsets that exhibited correlations in the same direction  (see
\textit{Exploratory correlation analyses}).  When all 10,000 randomly
chosen subsets of participants exhibited correlations in the same
direction (i.e., all positive correlations or all negative
correlations), we report the $p$-value as $p < 0.0001$.

@@

\section*{Discussion}
After collecting a year's worth of fitness-tracking data from each of 113
participants, we ran each participant in a battery of memory tasks and
had them fill out a series of demographic and mental health-related
questions.  We found that the associations between fitness-related
activities, memory performance, and mental health are complex.  For example,
participants who tended to engage in a particular intensity of physical activity
also tended to perform better on some memory tasks but worse on others.  This
suggests that engaging in one form or intensity of physical activity
will not necessarily affect all aspects of cognitive or mental health equally
(or in the same direction).

A number of prior studies have shown that engaging in exercise can improve cognitive
and mental health~\citep{ChanEtal12b, BrisEtal02, EtniEtal06, BassSuzu17, Ragl90, MikkEtal17, TaylEtal85, DeslEtal09, Call04,
PaluSchw00, MorrEtal18, GordEtal17, MorrEtal22, HerrEtal10}.  The majority of these studies
ask participants in an ``exercise intervention'' condition (where participants engage in
a designated physical activity or training regimen) or a ``control'' condition (where participants 
do not engage in the designated activity or training) to perform cognitive tasks or undergo
mental health screening.  In other words, most primary studies treat ``physical activity'' as a binary
variable that either is or is not present for each participant.  Most prior studies also 
track or manipulate exercise over relatively short durations (typically on the order of days or weeks).  Our current work indicates
that the true relations between physical activity, cognitive performance, and mental health
may be non-monotonic and heterogeneous across activities, tasks, and mental health measures.  These
relations can also unfold over much longer timescales than have been previously identified (on the order of months; Fig.~\ref{fig:dynamics}).
However, despite the complexities of the structures of these associations, we also found that 
they were often remarkably consistent across people.  For example, as displayed in Figure~\allCorr,
many of the associations between fitness, behavioral, and mental health
measures were consistent across over 95\% of 10,000 randomly chosen subsets
of participants.
